Introducing ICBe: An Event Extraction Dataset from Narratives about International Crises

### 2023-08-16 replication_paper/psrm_draft/ICBe_manuscript.Rmd
AG penultimate RR

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
authors:
  - name: Rex W. Douglass
    thanks: Correspondence should be addressed to Rex W. Douglass at rexdouglass@gmail.com.
    affiliation: University of California, San Diego
  - name: Thomas Leo Scherer
    affiliation: University of California, San Diego
  - name: J. Andrés Gannon
    affiliation: Vanderbilt University
  - name: Erik Gartzke
    affiliation: University of California, San Diego
  - name: Jon Lindsay
    affiliation: Georgia Institute of Technology
  - name: Shannon Carcelli
    affiliation: University of Maryland
  - name: Jonathan Wilkenfeld
    affiliation: University of Maryland
  - name: David M. Quinn
    affiliation: University of Maryland
  - name: Catherine Aiken
    affiliation: Georgetown University
  - name: Jose Miguel Cabezas Navarro
    affiliation: Universidad Mayor
  - name: Neil Lund
    affiliation: University of Maryland
  - name: Egle Murauskaite
    affiliation: University of Maryland
  - name: Diana Partridge
    affiliation: University of Maryland

@@

\doublespacing
\newpage

If we could record every international interaction in the realms of diplomacy, conflict, economics, and beyond, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? While this record could in principle be unbounded, the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be recorded and systematically measured by social scientists interested in these regularities.[^1] Thanks to improvements in natural language processing, more open-ended efforts have begun to capture entire unstructured streams of events from international news reports.[^2] This has invited fruitful efforts to evaluate the coverage, quality, and accuracy of attempts to measure international affairs.

@@

Much like the recording of chess evolved from natural language descriptive text notation to the modern figurine algebra notion, international relations scholars have recently sought to produce a structured account of historical events by combining the unstructured corpus of historical records with informative priors about international relations. The resulting structured account of information about a historical episode can be combined with that of other events to produce a systematic account of political events from which we can garner novel insight, assuming the underlying data have high coverage, precision, and recall. The easiest way to convey the desired produced of this task is with an example. @fig-cuba-narrative shows a narrative account of the Cuban Missile Crisis (1962) in natural language sentences alongside a mapping to discrete machine-readable abstractive events. From the structured data, scholars can identify similarities and differences across events concerning important concepts like when particular foreign policy actions deter versus inflame [@jervisCooperationSecurityDilemma1978; @glaserCausesConsequencesArms2000], when third parties mediate in interstate disputes [@haffarEmergentPeacemakersCataloguing2002; @quinnPowerPlayMediation2006], and how actors try to communicate resolve [@tragerDiplomacyWarPeace2016; @luptonReexaminingReputationResolve2018]. Identifying patterns of international interactions is not just an inherently interesting enterprise; it is a necessary precondition to important efforts to predict where policymakers should turn their attention to improve global welfare [@wardLearningSteppingFuture2013; @begerReassessingRoleTheory2021].

@@

![Cuban Missile Crisis (1962) - ICB Narrative vs. ICBe Events](case_study_cuban_precision.png){#fig-cuba-narrative width="7in" height="8in"}

@@

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].[^5] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (GUI) (SI Appendix 2.1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players, whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more specific and non-mutually exclusive activities. Some additional details were always collected (e.g. location and timing) while other details were only collected if appropriate (e.g. force size, fatalities, domains, units). We find `r codings_long_agreement %>% filter(varname_normalized %>% str_detect("actor")) %>% pull(value_normalized) %>% unique() %>% length()` actors and 117 different behaviors.[^6] A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions.

@@

Each crisis was typically assigned to 2 expert coders and 2 novice coders with an additional tie-breaking expert coder assigned to sentences with high disagreement.[^7] For the purposes of measuring intercoder agreement and consensus, we temporarily disaggregate the unit of analysis to the Coder-Crisis-Sentence-Tag (n=`r icb_long_clean %>% nrow() %>% scales::comma()`), where a tag is any unique piece of information a coder can associate with a sentence such as an actor, date, behavior, etc. We then aggregate those tags into final events (n=`r ICBe_events_agreed %>% nrow() %>% scales::comma()`), using a consensus procedure (SI Appendix 2.2) that requires a tag to have been chosen by at least one expert coder and either a majority of expert or novice coders. This screens noisy tags that no expert considered possible but leverages novice knowledge to tie-break between equally plausible tags chosen by experts.

[^7]: Expert coders were graduate students or postgraduates who collaboratively developed the ontology and documentation for the codebook. Undergraduate coders were students who engaged in classroom workshops.

@@

Our strategy for dealing with both problems is a plausibly objective yardstick called a synthetic historical narrative. We collect a large diverse corpus of narratives spanning timelines, encyclopedia entries, journal articles, news reports, websites, and government documents. Using natural language processing (fully described in SI Appendix 3.1), we identify details that appear across multiple accounts. The more accounts that mention a detail, the more central it is to understanding the true historical episode. The theoretical motivation is that authors face word limits which force them to pick and choose which details to include, and they choose details that serve the specific context of the document they are producing. With a sufficiently large and diverse corpus of documents, we can vary the context while holding the overall episode constant and see which details tend to be invariant to context. Intuitively, a high-quality event dataset should have high recall for context invariant details both because of their broader relevance and also because they are easier to find in source material.

We find substantive variation in recall across existing state of the art methods. Mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details.[^8] Furthermore, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events. @fig-recall-cases compares ICBe's recall with that of existing datasets for the two case studies detailed in Section 4. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Across all existing datasets and ICBe, recall increases with the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which in the Cuban Missile Crisis case is so noisy that it's recall curve is flat to decreasing as mentions increase. The two episode-level datasets (MIDs and ICM) have low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have higher coverage, with ICEWs outperforming Terrier. Importantly our corpus of ICB narratives has high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left little additional information on the table.

[^8]: As the ICB narratives are intended to explain conflictual behavior in a political context, many of the missing events concern more economic components of conflict (eg. nationalizing a foreign business). Even when they occur in the context of a crisis, these events largely fall outside the sample of information on which ICBe's ontology is currently trained. Even with this limitation, ICBe is more comprehensive than the existing datasets that do try to code the economic dimensions of these crises. We see expanding the ontology to broader international phenomenon as a promising future implementation of our model.

@@

How does ICBe's precision compare to the existing state of the art? A researcher should be able to lay out the events of a crisis on a timeline and read off the macrostructure of an episode from each individual move. We call this visualization a crisis map, a directed graph intersected with a timeline. A crisis map using ICBe for the Cuban Missile Crisis case study is provided in @fig-cuba-crisismap, and crisis maps for the two case studies using existing event datasets can be found in SI Appendix 4.3 and 4.4 and crisis maps for all crises using all datasets can be found on the companion website. The crisis maps reveal the episode-level datasets like MIDs or the original ICB are too sparse and vague to reconstruct the structure of the crisis (SI Appendix 4.3 and 4.4). On the other end of the spectrum, the high recall dictionary-based event datasets like Terrier and ICEWs produce so many noisy events (several hundred thousand) that even with heavy filtering their crisis maps are completely unintelligible. Further, because of copyright issues, none of these datasets directly provide the original text spans making event-level precision difficult to verify.

@@

@fig-umap shows the location of every sentence from the ICBe corpus in semantic space as embedded using the same large language model as before, and the median location of each ICBe event tag applied to those sentences.[^9] Labels reflect the individual leaves of the ontology and colors reflect the higher level coarse branch nodes of the ontology. If ICBe has high precision, substantively similar tags ought to have been applied to substantively similar source text, which is what we see both in two dimensions in the main plot and via hierarchical clustering on all dimensions in the dendrogram along the right-hand side.[^10]

[^9]: We preprocess sentences to replace named entities with a generic Entity token.

[^10]: Hierarchical clustering on cosine similarity and with Ward's method.

@@

In this section, we focus our validation on two case studies for which we have produced synthetic narratives using the method described in Section 3.2. Our proposed measure is a reconstruction task to see whether our intended ontology can be recovered through only unsupervised clustering of sentences they were applied to. The first (@fig-cuba-narrative) is the Cuban Missile Crisis (hereafter Cuban Missiles) which took place primarily in the second half of 1962, involved the United States, the Soviet Union, and Cuba, and is widely known for bringing the world to the brink of nuclear war. The second (SI Appendix 4.1) is the Crimea-Donbas Crisis (hereafter Crimea-Donbas) which took place primarily in 2014, involved Russia, Ukraine, and NATO, and within a decade spiraled into a full-scale invasion. We choose these cases because they are significant in contemporary international relations, are widely known across academic disciplines as well as among the public, and are sufficiently brief to evaluate in depth. They are similar in that both cases involve a superpower in crisis with a neighbor that changed from a friendly to a hostile regime, both held implications for the economic and military security for the superpower by risking full-scale invasion, and both eventually invited intervention by an opposing superpower.

@@

A synthetic historical narrative for Cuban Missiles appears in @fig-recall-cuba, with 51 events drawn from 2,020 documents. Each row represents a detail that appeared in at least five documents along with an approximate start date, a handwritten summary, the number of documents it was mentioned in, and whether it could be identified in the text of the original ICB corpus, our ICBe events, and any of the competing existing models.

ICB's improved recall of Cuban Missiles was relative to the state of the art was summarized in Section 3.2 (@fig-recall-cases), but the events that explain that improvement can now be seen. Our ground truth ICB narrative contains 17/51 of the events from the synthetic narrative of a case that includes high-level previously classified details. ICBe captures nearly all details included in ICB as well as more details from the synthetic narrative than any competing dataset. Phoenix includes some earlier information than ICBe like the nationalization of businesses and back channel negotiations, but the crisis narrative has a clean canonical end with the Soviets agreeing to withdraw missiles. ICBe stands out in including more communicative behavior (do -- speech) than existing datasets like US threats to attack and later promises not to invade. Given the recognized importance of threat credibility for understanding international conflict, the addition of this information is a substantively important improvement over the existing state of the art [@slantchevMilitaryThreatsCosts2011].

@@

@fig-cuba-crisismap shows the crisis map for the Cuban Missile Crisis. Looking at the crisis on a timeline, one can now identify the structure of actors and the environment, along with its supporting details, in a way that validates the precision of ICBe. Although harder to measure objectively, this crisis map provides face valid evidence that ICBe's account is not too vague, but also not unnecessarily detailed. We include much of the geopolitically important details like Soviet deployment, US discovery of that deployment, heightened alert levels, a blockade, and negotiations that ended with a formal agreement. At the same time, the crisis map indicates that ICBe does not include unnecessary nuances that preclude useful comparison to other international events.

@@

A synthetic historical narrative for the 2014 Crimea-Donbas crisis (30 events drawn from 971 documents) appears in @fig-recall-crimea. As in the earlier case, rows represent details that appeared in at least five documents and whether it is identified in ICBe and existing datasets.

Again quantitatively summarized earlier in Section 3.2 (@fig-recall-cases), our ground truth ICB narrative contains 23/30 of the events from the synthetic narrative. Like the gray zone precursor to the Cuban Missile crisis [@cormacGreyNewBlack2018], Ukraine provided several security guarantees to Russia that were potentially undone, e.g. a long term lease on naval facilities in Crimea. But unlike the Cuban Missile crisis, the end of this crisis is unclear, with the event meekly ending with a second cease-fire agreement (Minsk II) but continued fighting. ICBe again recalls more important information about the crisis than any existing dataset, particularly information concerning the behavior of non-state separatist groups like the Donetsk People's Republic (DPR) and Luhansk People's Republic (LPR).

@@

We take the same approach here in @fig-precision-icews, selecting four specific CAMEO event codings and checking how often they reflect a true real-world event from the Crimea-Donbas synthetic narrative. We choose four event types around key moments in the crisis. The start of the crisis revolves around Ukraine backing out of a trade deal with the EU in favor of Russia, but "sign formal agreement" events act more like a topic detector with dozens of events generated by discussions of a possible agreement but not the actual agreement which never materialized. The switch is caught by the "reject plan, agreement to settle dispute", but also continues for Viktor Yanukovych even after he was removed from power because of articles retroactively discussing the cause of his removal. Events for "use conventional military force" capture a threshold around the start of hostilities and who the participants were but not any particular battles or campaigns. Likewise, "impose embargo, boycott, or sanctions" captures the start of waves of sanctions and from who but are effectively constant as the news coverage does not distinguish between subtle changes or additions. In sum, dictionary-based methods on news corpora tend to have high recall because they parse everything in the news, but for the same reason, their specificity for most event types is too low to back out individual chess-like sequencing that ICBe aims to record.

@@

We release several open-source products along with supporting code and documentation to further advance the study of international relations, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).[^11] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here at crisisevents.org.

[^11]: Using ICBe data @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.

>>>>

# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
authors:
  - name: Rex W. Douglass
    thanks: Correspondence should be addressed to Rex W. Douglass at rexdouglass@gmail.com.
    affiliation: University of California, San Diego
  - name: Thomas Leo Scherer
    affiliation: University of California, San Diego
  - name: J Andrés Gannon
    affiliation: Vanderbilt University
  - name: Erik Gartzke
    affiliation: University of California, San Diego
  - name: Jon Lindsay
    affiliation: Georgia Institute of Technology
  - name: Shannon Carcelli
    affiliation: University of Maryland
  - name: Jonathan Wilkenfeld
    affiliation: University of Maryland
  - name: David M. Quinn
    affiliation: University of Maryland
  - name: Catherine Aiken
    affiliation: Georgetown University
  - name: Jose Miguel Cabezas Navarro
    affiliation: Universidad Mayor
  - name: Neil Lund
    affiliation: University of Maryland
  - name: Egle Murauskaite
    affiliation: University of Maryland
  - name: Diana Partridge
    affiliation: University of Maryland

@@

```{=tex}
\doublespacing
\newpage
```
If we could record every international interaction in the realms of diplomacy, conflict, economics, and beyond, how much unique information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? While this record could in principle be unbounded, the central conceit of social science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be recorded and systematically measured by social scientists interested in these regularities.[^1] Thanks to improvements in natural language processing, more open-ended efforts have begun to capture entire unstructured streams of events from international news reports.[^2] This has invited fruitful efforts to evaluate the coverage, quality, and accuracy of attempts to measure international affairs.

@@

Much like the recording of chess evolved from natural language descriptive text notation to the modern figurine algebra notion, international relations scholars have recently sought to produce a structured account of historical events by combining the unstructured corpus of historical records with informative priors about international relations. The resulting structured account of information about a historical episode can be combined with that of other events to produce a systematic account of political events from which we can garner novel insight, assuming the underlying data have high coverage, precision, and recall. The easiest way to convey the desired produced of this task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) in natural language sentences alongside a mapping to discrete machine-readable abstractive events. From the structured data, scholars can identify similarities and differences across events concerning important concepts like when particular foreign policy actions deter versus inflame [@jervisCooperationSecurityDilemma1978; @glaserCausesConsequencesArms2000], when third parties mediate in interstate disputes [@haffarEmergentPeacemakersCataloguing2002; @quinnPowerPlayMediation2006], and how actors try to communicate resolve [@tragerDiplomacyWarPeace2016; @luptonReexaminingReputationResolve2018]. Identifying patterns of international interactions is not just an inherently interesting enterprise; it is a necessary precondition to important efforts to predict where policymakers should turn their attention to improve global welfare [@wardLearningSteppingFuture2013; @begerReassessingRoleTheory2021].

@@

![Cuban Missile Crisis (1962) - ICB Narrative vs. ICBe Events](case_study_cuban_precision.png){#fig-cubanarrative width="7in" height="8in"}

@@

The ICBe ontology follows a hierarchical design philosophy where a smaller number of significant decisions are made early on and then progressively refined into more specific details [@brustIntegratingDomainKnowledge2020].[^5] Each coder was instructed to first thoroughly read the full crisis narrative and then presented with a custom graphical user interface (GUI) (SI Appendix 2.1). Coders then proceeded sentence by sentence, choosing the number of events (0-3) that occurred, the highest behavior (thought, speech, or activity), a set of players, whether the means were primarily armed or unarmed, whether there was an increase or decrease in aggression (uncooperative/escalating or cooperative/de-escalating), and finally one or more specific and non-mutually exclusive activities. Some additional details were always collected (e.g. location and timing) while other details were only collected if appropriate (e.g. force size, fatalities, domains, units). We find `r codings_long_agreement %>% filter(varname_normalized %>% str_detect("actor")) %>% pull(value_normalized) %>% unique() %>% length()` actors and 117 different behaviors.[^6]

@@

A unique feature of the ontology is that thought, speech, and do behaviors can be nested into combinations, e.g. an offer for the U.S.S.R. to remove missiles from Cuba in exchange for the U.S. removing missiles from Turkey. Through compounding, the ontology can capture what players were said to have known, learned, or said about other specific fully described actions. Two other novel features of our data -- its re-programmamable coding and its inclusion of thought, speech, and action -- intersect to provide a unique contribution. No existing event data distinguishes thoughts, speeches, and actions and most are limited to just coding actions given they are the most readily observable and thus the full universe could at least theoretically be included.[^7] Yet, methods like archival work and expert interviews continue to improve identification of thoughts and speeches that help us understand the reasoning process underlying political decisions and what actually occurred behind the scenes. This is an improvement over the state of the art that omits these by design in allowing scholars to identify thoughts associated with things like crisis escalation as well as patterns of omission to be advanced in future work, but the modular nature of our coding means the inclusion of a non-random sample of thoughts and speeches is neither necessary nor does it have inevitable downstream effects on the rest of the coding in ways that produce unwanted bias.

[^7]: Even the coding of overt actions like MIDs is not without contention [@giblerInternationalConflicts181620102018].

Each crisis was typically assigned to 2 expert coders and 2 novice coders with an additional tie-breaking expert coder assigned to sentences with high disagreement.[^8] For the purposes of measuring intercoder agreement and consensus, we temporarily disaggregate the unit of analysis to the Coder-Crisis-Sentence-Tag (n=`r icb_long_clean %>% nrow() %>% scales::comma()`), where a tag is any unique piece of information a coder can associate with a sentence such as an actor, date, behavior, etc. We then aggregate those tags into final events (n=`r ICBe_events_agreed %>% nrow() %>% scales::comma()`), using a consensus procedure (SI Appendix 2.2) that requires a tag to have been chosen by at least one expert coder and either a majority of expert or novice coders. This screens noisy tags that no expert considered possible but leverages novice knowledge to tie-break between equally plausible tags chosen by experts.

[^8]: Expert coders were graduate students or postgraduates who collaboratively developed the ontology and documentation for the codebook. Undergraduate coders were students who engaged in classroom workshops.

@@

Our strategy for dealing with both problems is a plausibly objective yardstick called a synthetic historical narrative. We collect a large diverse corpus of narratives spanning timelines, encyclopedia entries, journal articles, news reports, websites, and government documents. Using natural language processing (fully described in SI Appendix 3.1), we identify details that appear across multiple accounts. The more accounts that mention a detail, the more central it is to understanding the true historical episode. The theoretical motivation is that authors face word limits which force them to pick and choose which details to include, and they choose details that serve the specific context of the document they are producing. With a sufficiently large and diverse corpus of documents, we can vary the context while holding the overall episode constant and see which details tend to be invariant to context. This presents a harder evaluation baseline than comparing ICBe's recall to just that of ICB since there are non-crisis aspects of these events that may be included in other narratives but are out of the scope of our data. For example, the nationalization of businesses in Cuba may be included as important context in the Cuban Missile Crisis in documents that do not focus on the crisis dimensions like ICB. Using this hard case, a recall measure of ICBe on the synthetic narratives thus serves as a way to evaluate the breadth of ICBe's ontology and potential application to non-crisis international events.

We find substantive variation in recall across existing state of the art methods. Mentions of a detail across accounts are exponentially distributed with context-invariant details appearing dozens to hundreds of times more than context-dependent details.[^9] Furthermore, crisis start and stop dates are arbitrary, and the historical record points to many precursor events as necessary detail for understanding later events. Figure 2 compares ICBe's recall with that of existing datasets for the two case studies detailed in Section 4. ICBe strictly dominates all of the systems but ICEWs in recall though we note that the small sample sizes mean these systems should be considered statistically indistinguishable. Across all existing datasets and ICBe, recall increases with the number of document mentions which is an important sign of validity for both them and our benchmark. The one outlier is Phoenix which in the Cuban Missile Crisis case is so noisy that it's recall curve is flat to decreasing as mentions increase. The two episode-level datasets (MIDs and ICM) have low coverage of contextual details. The two other dictionary systems ICEWs and Terrier have higher coverage, with ICEWs outperforming Terrier. Importantly our corpus of ICB narratives has high recall of frequently mentioned details giving us confidence in how those summaries were constructed, and ICBe lags only slightly behind showing that it left little additional information on the table.

[^9]: As the ICB narratives are intended to explain conflictual behavior in a political context, many of the missing events concern more economic components of conflict (eg. nationalizing a foreign business). Even when they occur in the context of a crisis, these events largely fall outside the sample of information on which ICBe's ontology is currently trained. Even with this limitation, ICBe is more comprehensive than the existing datasets that do try to code the economic dimensions of these crises. We see expanding the ontology to broader international phenomenon as a promising future implementation of our model.

@@

How does ICBe's precision compare to the existing state of the art? A researcher should be able to lay out the events of a crisis on a timeline and read off the macrostructure of an episode from each individual move. We call this visualization a crisis map, a directed graph intersected with a timeline. A crisis map using ICBe for the Cuban Missile Crisis case study is provided in Figure 2, and crisis maps for the two case studies using existing event datasets can be found in SI Appendix 4.3 and 4.4 and crisis maps for all crises using all datasets can be found on the companion website. The crisis maps reveal the episode-level datasets like MIDs or the original ICB are too sparse and vague to reconstruct the structure of the crisis (SI Appendix 4.3 and 4.4). On the other end of the spectrum, the high recall dictionary-based event datasets like Terrier and ICEWs produce so many noisy events (several hundred thousand) that even with heavy filtering their crisis maps are completely unintelligible. Further, because of copyright issues, none of these datasets directly provide the original text spans making event-level precision difficult to verify.

@@

Figure 3 shows the location of every sentence from the ICBe corpus in semantic space as embedded using the same large language model as before, and the median location of each ICBe event tag applied to those sentences.[^10] Labels reflect the individual leaves of the ontology and colors reflect the higher level coarse branch nodes of the ontology. If ICBe has high precision, substantively similar tags ought to have been applied to substantively similar source text, which is what we see both in two dimensions in the main plot and via hierarchical clustering on all dimensions in the dendrogram along the right-hand side.[^11]

[^10]: We preprocess sentences to replace named entities with a generic Entity token.

[^11]: Hierarchical clustering on cosine similarity and with Ward's method.

@@

In this section, we focus our validation on two case studies for which we have produced synthetic narratives using the method described in Section 3.2. Our proposed measure is a reconstruction task to see whether our intended ontology can be recovered through only unsupervised clustering of sentences they were applied to. The first (Figure 1) is the Cuban Missile Crisis (hereafter Cuban Missiles) which took place primarily in the second half of 1962, involved the United States, the Soviet Union, and Cuba, and is widely known for bringing the world to the brink of nuclear war. The second (SI Appendix 4.1) is the Crimea-Donbas Crisis (hereafter Crimea-Donbas) which took place primarily in 2014, involved Russia, Ukraine, and NATO, and within a decade spiraled into a full-scale invasion. We choose these cases because they are significant in contemporary international relations, are widely known across academic disciplines as well as among the public, and are sufficiently brief to evaluate in depth. They are similar in that both cases involve a superpower in crisis with a neighbor that changed from a friendly to a hostile regime, both held implications for the economic and military security for the superpower by risking full-scale invasion, and both eventually invited intervention by an opposing superpower.

@@

A synthetic historical narrative for Cuban Missiles appears in Figure 4, with 51 events drawn from 2,020 documents. Each row represents a detail that appeared in at least five documents along with an approximate start date, a handwritten summary, the number of documents it was mentioned in, and whether it could be identified in the text of the original ICB corpus, our ICBe events, and any of the competing existing models.

ICB's improved recall of Cuban Missiles was relative to the state of the art was summarized in Section 3.2 (Figure 2), but the events that explain that improvement can now be seen. Our ground truth ICB narrative contains 17/51 of the events from the synthetic narrative of a case that includes high-level previously classified details. ICBe captures nearly all details included in ICB as well as more details from the synthetic narrative than any competing dataset. Phoenix includes some earlier information than ICBe like the nationalization of businesses and back channel negotiations, but the crisis narrative has a clean canonical end with the Soviets agreeing to withdraw missiles. ICBe stands out in including more communicative behavior (do -- speech) than existing datasets like US threats to attack and later promises not to invade. Given the recognized importance of threat credibility for understanding international conflict, the addition of this information is a substantively important improvement over the existing state of the art [@slantchevMilitaryThreatsCosts2011].

@@

Figure 5 shows the crisis map for the Cuban Missile Crisis. Looking at the crisis on a timeline, one can now identify the structure of actors and the environment, along with its supporting details, in a way that validates the precision of ICBe. Although harder to measure objectively, this crisis map provides face valid evidence that ICBe's account is not too vague, but also not unnecessarily detailed. We include much of the geopolitically important details like Soviet deployment, US discovery of that deployment, heightened alert levels, a blockade, and negotiations that ended with a formal agreement. At the same time, the crisis map indicates that ICBe does not include unnecessary nuances that preclude useful comparison to other international events.

@@

A synthetic historical narrative for the 2014 Crimea-Donbas crisis (30 events drawn from 971 documents) appears in Figure 6. As in the earlier case, rows represent details that appeared in at least five documents and whether it is identified in ICBe and existing datasets.

Again quantitatively summarized earlier in Section 3.2 (Figure 2), our ground truth ICB narrative contains 23/30 of the events from the synthetic narrative. Like the gray zone precursor to the Cuban Missile crisis [@cormacGreyNewBlack2018], Ukraine provided several security guarantees to Russia that were potentially undone, e.g. a long term lease on naval facilities in Crimea. But unlike the Cuban Missile crisis, the end of this crisis is unclear, with the event meekly ending with a second cease-fire agreement (Minsk II) but continued fighting. ICBe again recalls more important information about the crisis than any existing dataset, particularly information concerning the behavior of non-state separatist groups like the Donetsk People's Republic (DPR) and Luhansk People's Republic (LPR).

@@

We take the same approach here in Figure 7, selecting four specific CAMEO event codings and checking how often they reflect a true real-world event from the Crimea-Donbas synthetic narrative. We choose four event types around key moments in the crisis. The start of the crisis revolves around Ukraine backing out of a trade deal with the EU in favor of Russia, but "sign formal agreement" events act more like a topic detector with dozens of events generated by discussions of a possible agreement but not the actual agreement which never materialized. The switch is caught by the "reject plan, agreement to settle dispute", but also continues for Viktor Yanukovych even after he was removed from power because of articles retroactively discussing the cause of his removal. Events for "use conventional military force" capture a threshold around the start of hostilities and who the participants were but not any particular battles or campaigns. Likewise, "impose embargo, boycott, or sanctions" captures the start of waves of sanctions and from who but are effectively constant as the news coverage does not distinguish between subtle changes or additions. In sum, dictionary-based methods on news corpora tend to have high recall because they parse everything in the news, but for the same reason, their specificity for most event types is too low to back out individual chess-like sequencing that ICBe aims to record.

@@

We release several open-source products along with supporting code and documentation to further advance the study of international relations, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).[^12] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here at crisisevents.org.

[^12]: Using ICBe data, @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.
